---
title: "558_Project_3"
author: "ankit gupta & brian higginbotham"
date: "2023-11-08"
output:
  github_document:
    html_preview: false
params:
  education: "Elementary"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error = FALSE)
```

```{r parameter, echo=FALSE}
params$education
```

# Introduction

# Data

For the longer categorical variables - education, age, and income - the initial data set did not provide a thorough description of the categories. References to [this codebook](https://www.cdc.gov/brfss/annual_data/2015/pdf/codebook15_llcp.pdf) were used for the full descriptions of the categorical variables.

In order to conduct our analysis on separate education levels, we'll need do so some data prep on the original file. First, the education variable will need to be reduced to five categories by combining the first and second levels (the "no education" and "elementary education" levels.) We will then convert the variable to a factor and add the appropriate level descriptions that will be used in the report automation.
```{r data-read, results = "hide"}
library(tidyverse)
library(dplyr)
data <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
diabetes <- data %>% mutate(Education = if_else(Education <= 2, 1, 
                    if_else(Education == 3, 2, if_else(Education == 4, 3, 
                    if_else(Education == 5, 4, 5))))) %>%
  mutate(Education = factor(Education)) %>%
  mutate(Education = recode(Education, "1" = "Elementary", "2" = "HighSchool", 
                  "3" = "HighSchoolGrad", "4" = "College", "5" = "CollegeGrad")) %>% 
# This is where automation will select the correct subset
  filter(Education == params$education) %>%
  dplyr::select(-Education)
```
Most columns in the initial dataset are binary (0 or 1) for the variable description but a few represent categorical data coded numerically and a few also represent actual numerical measurements. We'll  convert the categorical variables to factors to help with the Exploratory Data Analysis as well as a few binary variables. Again, the descriptions for the categories was taken from the above codebook.
```{r factoring}
### MentHlth & PhysHlth converted to binary
diabetes <- diabetes %>% 
  mutate(MentHlth = if_else(MentHlth == 0, 0, 1)) %>%
  mutate(PhysHlth = if_else(PhysHlth == 0, 0, 1)) %>%
### create factors
  mutate(GenHlth = factor(GenHlth)) %>%
  mutate(GenHlth = recode(GenHlth, "1" = "excellent", "2" = "very_good", 
                          "3" = "good", "4" = "fair", "5" = "poor")) %>%
  mutate(Sex = factor(Sex)) %>%
  mutate(Sex = recode(Sex, "0" = "female", "1" = "male")) %>%
  mutate(Age = factor(Age)) %>%
  mutate(Age = recode(Age, "1" = "18_24", "2" = "25_29", "3" = "30_34", 
                      "4" = "35_39", "5" = "40_44", "6" = "45_49", "7" = "50_54", 
                      "8" = "55_59", "9" = "60_64", "10" = "65_69", "11" = "70_74", 
                      "12" = "75_79", "13" = "> 80")) %>%
  mutate(Income = factor(Income)) %>%
  mutate(Income = recode(Income, "1" = "10k", "2" = "15k", "3" = "20k", "4" = "25k", 
                         "5" = "35k", "6" = "50k", "7" = "75k", "8" = "greater_75k")) %>%
  mutate(Smoker = factor(Smoker)) %>%
  mutate(Smoker = recode(Smoker, "0" = "no", "1" = "yes"))
```
Two variables were re-coded with respect to modeling. `MentHlth` and `PhysHlth` were both measurements of "number of days of poor mental or physical health within the last thirty days." We can see from the data description provided by the "Diabetes Health Indicators Dataset" that the distribution of both variables is highly skewed towards 0 - no reported days. We decided to convert both variables to binary - 0 for "no reported days" and 1 to indicate "at least one reported day."

We will wait to convert our response variable - `Diabetes-binary` - to a factor after we complete EDA. We want to keep this variable as numeric so we can run quantitative summaries on it.

# Summarizations

First, let's get an idea of the overall diabetes rate for this education level.
```{r mean_diabetes}
mean(diabetes$Diabetes_binary)
```
How does the rate change with age?
```{r mean_by_age}
diabetes %>% group_by(Age) %>% summarize(mean(Diabetes_binary), sd(Diabetes_binary))
```
Is there a prevelance by gender?
```{r contigency_sex}
table(diabetes$Sex, diabetes$Diabetes_binary)
```
Does a persons income level have any influence on diagnosis? If not, then we would expect the diagnosis rate of people with diabetes to be about the same across income levels.
```{r mean_by_income}
diabetes %>% group_by(Income) %>% summarize(mean(Diabetes_binary), sd(Diabetes_binary))
```

How does BMI relate to diabetes? Here we'll see a density histogram that shows the distribution of both diabetes (1) and no diabetes (0) by BMI.
```{r BMI_hist}
library(ggplot2)
g <- ggplot(diabetes, aes(x=BMI))
g + geom_histogram(aes(fill = as.factor(Diabetes_binary), y = ..density..), position = "dodge") +
  labs(title = "Density Histogram - BMI", fill = "Diabetes")
```
<br>

From the data description, the `GenHlth` variable appears to be self-reported. It may be interesting to see how the BMI compares to `GenHlth`. Of course, there are many variables that contribute to an assessment of "General Health," and `BMI` is not necessarily an accurate indicator of "General Health," but an additional layer for `Diabetes_binary` may gives us some insight if `GenHlth` could be a good variable in predicting diabetes.
```{r bmi_genhlth}
g <- ggplot(diabetes, aes(x = GenHlth, y = BMI))
g + geom_point(aes(color = as.factor(Diabetes_binary)), position = "jitter") +
  labs(y = "BMI", title = "Self-Reported Health vs. BMI", x = "General Health") + 
  guides(color = guide_legend(title = "diabetes"))
```
<br>

It's likely some of the variables will be correlated since so many are health indicators, so here's a correlation plot that will show us the strength of any correlations.
```{r corrplot}
library(corrplot)
correlation <- cor(dplyr::select(diabetes, Diabetes_binary, HighBP, HighChol, CholCheck, Stroke, HeartDiseaseorAttack, PhysActivity, Fruits, Veggies, HvyAlcoholConsump, AnyHealthcare, NoDocbcCost, DiffWalk, MentHlth, PhysHlth), method = "spearman")
corrplot(correlation, type = 'upper', tl.pos = 'lt')
```
<br>





```{r vcd}

library(vcd)
mosaicplot(~ Income + GenHlth, data = diabetes, color = TRUE, main = 'General Health by Income Bracket')

```


Here we can see general trend where individuals with better health often have a higher income. There are more factors in play, but it is generally assumed that higher income individuals will be able to afford more services.


```{r stack}

library(vcd)
mosaicplot(~ Age + GenHlth, data = diabetes, color = TRUE, main = 'General Health by Age')

```
Another interesting finding here is that health starts to level out at Age of 50. Also intersting is that folks past the age of 80 reported improved health over the brackets before. Perhaps life satisfaction and more time outside of the workplace could be hypothetical explanations.


```{r stack}
library(ggplot2)

# Assuming mydata is your data frame
ggplot(diabetes, aes(x = GenHlth, y = BMI)) +
  geom_violin() +
  labs(title = "Violin Plot of BMI by Health Eval",
       x = "Health",
       y = "BMI") +
  theme_minimal()
```
Interesting here that there doesn't seem to be a great difference in BMI by way of Health which is head scratching. However, BMI is not the most accurate predictor of prime health, as muscle mass can skew these numbers.


# Modeling

Before we start modeling, we'll need to create a new factor column based on `Diabetes_binary`. This factored version will be used in modeling and the numeric version we'll save until the data split.
```{r data-split}
library(caret)
diabetes$Diabetes_fac <- if_else(diabetes$Diabetes_binary == 0, "negative", "positive")
diabetes$Diabetes_fac <- as.factor(diabetes$Diabetes_fac)

### Data Split
set.seed(21)
dfIndex <- createDataPartition(diabetes$Diabetes_fac, p = 0.7, list = FALSE)
diabetesTrain <- diabetes[dfIndex, ]
diabetesTest <- diabetes[-dfIndex, ]
# create a numeric vector for final model selection
test_logloss <- diabetesTest$Diabetes_binary
# delete the Diabetes_binary column from both test and train sets
diabetesTrain <- diabetesTrain %>% dplyr::select(-Diabetes_binary)
diabetesTest <- diabetesTest %>% dplyr::select(-Diabetes_binary)
```
At the end, we'll be selecting the best performing model based on `LogLoss` results. In order to run the `LogLoss()` function efficiently, we'll need a vector of the results from the test set. So we've created that vector as `test_logloss` and then deleted the numeric columns `Diabetes_binary` from both the test and train sets so they will not interfere with the modeling and testing.
<br>

## LogLoss

We're going to use LogLoss as the measurement for model performance. LogLoss measures the distance between correct outcome (1 or 0) and the calculated probability. So the closer the model prediction is to the actual outcome, the lower the LogLoss score is. 

LogLoss can be a better way of measuring the performance as compared to an Accuracy rating. If a model predicted 0.51 probability as True and the actual result was True, this would be marked as a correct outcome in an accuracy rating. But in reality, the model barely got the prediction right and this may result in poor performance when the model is implemented on new data. LogLoss records this measured discrepancy and thus may provide a better description of the models performance. LogLoss is similar to the Residual Mean Squared Error (RMSE) produced in linear regression models.

## Logistic Regression

Logistic Regression is a Generalized Linear Model that models the probability of a binary outcome (success or failure, 1 or 0, etc). A Simple Linear Regression model would produce a continuous response on the real line that would not correspond to the predictions we are trying to make, so we use a logistic function that produces a logit or log-odds of success that is linear in its parameters. The logit or log-odds is then used to determine the outcome, generally split at probability 0.5 (greater than 0.5 = success, less than 0.5 = failure).

We'll fit three different Logistic Regression models and evaluate the training results to select the best performing model. The first model will train on all variables. The second will combine some main effects plus a polynomial term. The third will look at some more main effects and also include some interactions.
```{r log_reg}
set.seed(223)
logreg_fit1 <- train(Diabetes_fac ~ ., data=diabetesTrain, method="glm", family="binomial",
                     metric="logLoss",
                  preProcess=c("center", "scale"),
                  trControl = trainControl(method="cv", number = 5, 
                                           classProbs=TRUE, summaryFunction=mnLogLoss))

logreg_fit2 <- train(Diabetes_fac ~ BMI + poly(Smoker, 2, raw=TRUE) + Fruits:Veggies + HighBP + HighChol, data=diabetesTrain, method="glm", family="binomial",
                     metric="logLoss",
                     preProcess=c("center", "scale"),
                     trControl = trainControl(method="cv", number = 5, 
                                              classProbs=TRUE, summaryFunction=mnLogLoss))

logreg_fit3 <- train(Diabetes_fac ~ BMI + Age + Income + BMI:Age + BMI:Income + 
                       HighBP:HighChol + Smoker, data=diabetesTrain,
                     method="glm", family="binomial",
                     metric="logLoss",
                     preProcess=c("center", "scale"),
                     trControl = trainControl(method="cv", number = 5, 
                                              classProbs=TRUE, summaryFunction=mnLogLoss))

logreg_fit1$results
logreg_fit2$results
logreg_fit3$results
```
The best LogLoss value of the three models is `logreg_fit1`. 

##Lasso Regression

Lasso regression, short for Least Absolute Shrinkage and Selection Operator, is a linear regression technique that includes a penalty term to the ordinary least squares (OLS) objective function. The purpose of this penalty term is to encourage the model to prefer simpler models with fewer predictor variables, effectively driving some of the coefficients to zero.

In lasso regression, the objective function is modified by adding the sum of the absolute values of the regression coefficients multiplied by a regularization parameter (often denoted as λ

In classification tasks, logistic regression is a standard choice, while lasso regression is more commonly used for regression problems with a focus on regularization and feature selection.


```{r las_reg}
library(caret)
diabetesTest

log_lasso <- train(Diabetes_fac ~ . ,data=diabetesTrain,
                  method="glmnet",
                  family="binomial",
                  preProcess = c("center", "scale"),
                   trControl = trainControl(method="cv", number = 5, 
                                              classProbs=TRUE, summaryFunction=mnLogLoss))



log_lasso$results


#alpha of .10 and regularization parameter (lambda) of 0.0230916176 gave us the strongest model
                 
```


## Random Forest

Random Forest utilizes bootstrap aggregation to calculate the average response over many fitted trees. Whereas a Classification and Bagged Tree model will use all predictors in modeling, Random Forest will use a random subset of predictors for each bootstrap sample. By randomly selecting a subset of predictors, a good predictor or two will not dominate the tree fits and thus the variance from aggregation will not be reduced, resulting in a more robust model.

We'll go ahead and train a random forest fit on 1 to 6 predictors. The nice thing about the `caret::train` function is that it will automatically select the best fit to use when we run our predictions, so we will not need to evaluate mulitple results.


```{r random_forest}

rf_fit <- train(Diabetes_fac ~., data = diabetesTrain, method = "rf",
              metric = "logLoss",
              preProcess = c("center", "scale"),
              trControl = trainControl(method = "cv", number = 5, classProbs=TRUE, summaryFunction=mnLogLoss),
              tuneGrid = data.frame(mtry = 1:4))

rf_fit$results
rf_fit$bestTune
```
We went ahead and printed the results and the `$bestTune` to confirm that `caret::train` has selected the best model.

Setting mtry at 4 gave us the best intro sample to train the model. We made that the max due to computational effort.


## Support Vector Machine


A Support Vector Machine (SVM) is a supervised machine learning algorithm that can be used for both classification and regression tasks. However, it is more commonly associated with classification problems. The primary objective of an SVM is to find a hyperplane that best separates data points of different classes in a high-dimensional space. The hyperplane is chosen in such a way that it maximizes the margin between the two classes.

SVMs are effective in high-dimensional spaces and are widely used in various applications, including image classification, handwriting recognition, bioinformatics, and text classification. They are particularly useful when the data has complex relationships and when a clear margin of separation between classes is desirable. SVMs have both linear and non-linear variants, making them versatile for a range of problems.


```{r svm}


svm <- train(Diabetes_fac ~., data = diabetesTrain, method = "svmLinear", 
               trControl = trainControl(method = "cv", number = 5, classProbs=TRUE, summaryFunction=mnLogLoss),  
               preProcess = c("center","scale"))

svm
```
SVM is a viable model as it has an ogloss of .5344. The model does better with more complex data making it better with higher dimensional data,



## Linear Discriminant Analysis

Linear Discriminant Analysis identifies the most discriminitive features of the dataset to enhance classification accuracy. It projects the data onto a lower dimensional space (dimensionality reduction) and finds a linear combination of variables that best separates the classes in the dataset. One of it's advantages is that it can handle multicollineartity within the data. However, a few disadvantages are that it assumes a normal distribution and that the data is linearly separable - both of which may not always be the case.

We'll go ahead and fit a LDA model using all variables and compare the results to our previous models.
```{r lda}
library(MASS)
lda_fit <- train(Diabetes_fac ~ ., data = diabetesTrain, method = "lda",
                 metric = "logLoss",
                 PreProcess = c("center", "scale"),
                 trControl = trainControl(method = "cv", number = 5, 
                                          classProbs=TRUE, summaryFunction=mnLogLoss))

lda_fit$results
```
The logLoss for this model is better than the logistic regression and random forest. So we'll go ahead and select the fit using all variables.

# Final Model Selection

Now that we have our six trained models, we'll test them on the test data and select the model with the lowest logLoss as the winner. To do this, we'll write a function that will take in a list of the model names and run them through a `for loop`  that will run `predict()` and calculate `LogLoss()` on each with the results appended to a list. Note the `test_logloss` argument that we pulled from our test data before we deleted the `Diabetes_binary` columns. This `LogLoss()` function is what we were saving it for.
```{r model_selection}
library(MLmetrics)
model_list <- list(logreg_fit1, lda_fit)

best_model <- function(x){
  results <- list()
  for(i in 1:length(x)){
    pred <- data.frame(predict(x[i], newdata = diabetesTest, type = "prob"))
    l_loss <- LogLoss(pred$positive, test_logloss)
    results[i] <- l_loss
  }
  names(results) <- c("logreg", "linear discriminate")
  return(results)
}

logloss_results <- data.frame(best_model(model_list))
```
Now we have the LogLoss score for each of our models stored in `logloss_results`. Let's take a look at all the results.
```{r results}
logloss_results
```
The best model
```{r best, echo=FALSE}
which.min(logloss_results)
```